Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-09-29 SPH-Torus-paper1/sphtorus_benchmark.tex
Add section on transient oscillations.
Spell checked.


git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2037 25464f21-6099-402c-b30a-640f1d40d38b

@@

  We present results from combining a grid-based radiative transfer
  code with a Smoothed Particle Hydrodynamics code to produce a
  flexible system for modelling radiation hydrodynamics. We use a
  benchmark model of a circumstellar disc to determine a robust method
  for constructing a gridded density density distribution from SPH
  particles. The benchmark disc is then used to determine the accuracy
  of the radiative transfer results. We find that the SED and 
  the temperature distribution within the disc are sensitive
  to the representation of the disc inner edge, which depends
  critically on both the grid and SPH resolution. The code is then
  used to model a circumstellar disc around a T-Tauri star. As the
  disc adjusts towards equilibrium vertical motions in the disc are
  induced resulting in scale height enhancements which intecept
  radiation from the central star. Vertical transport of radiation
  enables these perturbations to influence the mid-plane temperature
  of the disc. The vertical motions decay over time and the disc
  ultimately reaches a state of simulatanous hydrostatic and radiative
  equilibrium.

@@

Accurately coupling the radiative transfer component to the SPH
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
cicumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

The accuracy of the SPH method has previously been examined in detail
(e.g. \cite{price_2008b} and references therein) so we do not attempt
to benchmark the SPH component of the code here. Radiative transfer in
a circumstellar disc has been studied in detail by
\cite{pascucci_2004} and this benchmark is regularly used to test the
accuracy of the {\small{TORUS}} code. Consequently we can be confident
in the accuracy of the SPH and raditive transfer codes individually so
we focus here on the accuracy of the mechansim by which they are
coupled. Hence the benchmark case presented here is a ``static''
benchmark in which we represent an analytical density distribution
with SPH particles and test whether the radiative transfer code can
calculate accurate results using this SPH density representation.

@@

An important step in setting up grid-based radiative transfer using a
density distribution derived from SPH particles is determining an
effective transformation from the Lagrangian particle description to
the Eulerian grid desciption.  The representation on grid should
preserve important properties of the SPH model (e.g. total mass) but
also needs to take account of the possibility that a radiative
transfer calculation requires the highest resolution in different
regions to a hydrodynamics calculation (e.g. to resolve opacity
gradients). This section determines a grid construction method which
allows a good degree of control over the number of cells in the grid,
and the location of the highest resolution regions, while maintaining
an accurate representation of the total disc mass.

@@

The density in a given grid cell is calculated using an exponential
kernel smooth. If the sum of the kernel weights is greater than 0.3
the density is normalised by the sum of the kernel weights, to ensure a
smooth density distribution within the interior of the disc. If the
sum of the weights is less than 0.3 then the density is not normalised
by the sum of the weights, in order to avoid numerical effects at the
free surface, as described by \cite{price_2007}.

%The method used to
%construct an AMR grid from SPH particles is described in more detail
%in \cite{rundle_2009}.

@@

For each splitting condition a number of AMR grids were generated
using different values of the mass per cell limit or
$f_{\rm{split}}$. This was repeated for discs represented by $10^5$,
$10^6$ and $10^7$ SPH particles. Figure~\ref{fig:grid_tests} plots the
number of octals generated as a function of mass per cell or
$f_{\rm{split}}$ (solid line), and the percentage error in the disc
mass (dashed line). The percentage error in disc mass is calculated by
comparing the total mass on the AMR grid with the known mass of
$0.011~\rm{M}_{\sun}$ based on the analytical form of the density
distribution, and a positive value indicates that mass on the grid is
too large.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for cartesian AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:grid_tests}
\end{figure*}

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but it needs to be modified to take into account the increase in 
cell volume as cylindrical polar radius increases, for fixed values of
$dr$ and $dz$.  
%
Consider a cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in the cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{az}}}
\end{equation}
if there are $n_{\rm{az}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{az}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{az}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

The grid generation tests were run using a cylindrical polar geometry
with the modified mass per cell limit described above and an
unmodified density contrast condition. The results are plotted in
Fig.~\ref{fig:cp_grid_tests}.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for cylindrical polar AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:cp_grid_tests}
\end{figure*}
Using a cylindrical polar geometry results in a significantly reduced
number of octals in the AMR grid but with larger mass errors. Provided
the mass per cell limit is no more than $10^{27}\rm{g}$ then the total
mass is well represented with significantly fewer cells than in a
cartesian grid. The density contrast condition alone does not
represent the total mass to within 3\% accuracy even with
$f_{\rm{split}}=0.1$. Although this condition successfully adds
enhanced resolution in regions of higher density contrast it needs to
be combined with a mass per cell limit to ensure a robust total mass
representation.

@@

Having investigated the procdure for setting up the radiative transfer
grid we now proceed to studying the accuracy of the results from the
radiative transfer calculation. This was carried out by comparing
temperature distributions and spectral energy distributions (SEDs)
from the benchmark disc. Temperature distribution errors provide a
clear picture of where errors in the radiative transfer calculation
are located whereas SEDs provide a direct link to an observable
quantity and are a stringent test of the accuracy of the radiative
transfer calculation.

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid contructed from a mass per cell limit of $5\times10^{26}$g.  The
mass per cell limit was chosen in preference to the density contrast
limit to give the highest resolution near the central region of the
disc, which we expect to be most important for determining the
SED. SEDs were calculated using the {\small{TORUS}} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc with $10^5 $ particles. SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right). The benchmark SED is plotted as a solid line and
    the {\small{TORUS}} SED is plotted as a dashed line.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1~$\mu\rm{m}$. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1AU gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1AU and secondly the smoothing lengths of the particles are larger
than 1AU. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in central region of disc with $10^5$ particles. Axis
units are $10^{13}$cm.}
\label{fig:basic_dens}
\end{figure}

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc
(i.e. the ``prompt escape'' regime decribed by \cite{nayakshin_2009}).
A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years). The freqency
with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc.

A cylindrical polar grid, with adaptive $r$ and $z$ cells and 16
azimuthal cells, was used for the raditive transfer step. The AMR grid
was constructed using a maximum mass per cell of
$3.0\times10^{25}~\rm{g}$ and a density splitting condition of
$f_{\rm{split}}=0.1$. The density in a cylindrical region around the
source was reduced to a very low value to ensure that the disc did not
completely obscure the source. A real circumstellar disc will be
truncated by the stellar magnetosphere so there is a physical
justification for enforcing a low density region around the star. The
radius of this region is $R_{\rm{gap}}=1.0~\rm{AU}$ which corresponds
to the accretion radius of the SPH sink particle. Two extra levels of
grid refinement were added in a box around the central source to
ensure that the grid had enough resolution to allow the gap around the
source to be resolved. The first box of extra refinement has size
$10^{14}~\rm{cm}$ with cells no larger than $10^{13}~\rm{cm}$ and the
second box has size $4\times10^{13}~\rm{cm}$ with cells no larger than
$4\times10^{12}~\rm{cm}$.

@@

Plots of the disc internal energy are shown in Fig.~\ref{fig:log_u} at
six different times during the simulation. Figure~\ref{fig:frame1}
shows the initial state before the radiative transfer is switched on
when there is a vertically isothermal equation of state. The second
frame (Fig.~\ref{fig:frame2} is shortly after the raditive transfer
has been switched on. The inner disc has collapsed and there is
heating seen in parts of the disc exposed to radiation from the
central star. In Fig.~\ref{fig:frame3} a scale height enhancement is
present in the disc at a radius of approximately
$3\times10^{15}~\rm{cm}$ which shadows the region immediately behind
causing it to cool. Figure~\ref{fig:frame4} shows several such scale
height enhancements and in each case there is cooling behind the
enhancement as the scale height fluctations are able to affect the
mid-plane temperature due to vertical transport of radiation. In
Fig.~\ref{fig:frame5} the inner region of the disc has settled and the
scale height enhancements are seen only in the outer disc and by the
last frame (Fig.~\ref{fig:frame6}) the disc has reached a steady state
in hydrostatic and radiative equilibrium.
\begin{figure*}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6675}\label{fig:frame1}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6691}\label{fig:frame2}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6763}\label{fig:frame3}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC7363}\label{fig:frame4}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC9363}\label{fig:frame5}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_run2_DSC1024}\label{fig:frame6}}
  \caption{Particle plots showing internal energy in cyclindrical
    polar co-ordinates. Axis units are cm.}
  \label{fig:log_u}
\end{figure*}

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SEDs. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a managable
level with a realistic number of SPH particles in the disc.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. As the disc settles towards an equilibrium state,
vertical motions are seen which give the appearance of an outward
propagating scale height fluctuation. This is a vertical motion which
decays more rapidly at smaller radii and is seen whether radiative
transfer is included or not. These are not the same as the
instabilities seen in other radiation hydrodynamics disc models
\citep{min_2009}. One consequence of these fluctuations is that they
can intecept radiation from the central star which can then be
vertically transported into the mid-plane of the disc hence even
thought the fluctuations are of a small amplitude they exert an
influence on the mid-plane temperature.

>>>>

\def\aap{{ A\&A}}
\def\AaA{{ A\&A}}
\def\aaps{{ A\&AS}}
\def\AJ{{AJ}}
\def\aj{{AJ}}
\def\ApJ{{ApJ}}
\def\apj{{ApJ}}
\def\apjl{{ApJL}}
\def\ApJL{{ApJL}}
\def\ApJSupp{{ApJS}}
\def\MN{{MNRAS}}
\def\mnras{{MNRAS}}
\def\pasj{{PASJ}}
\def\nat{Nature}
\def\ARAA{{ARA\&A}}
\def\Obs{{ Observatory}}
\def\PTRS{{Phil. Trans. Roy. Soc.}}
\def\azh{{Astronomicheskii Zhurnal}}
\def\apjs{{ApJS}}
\def\araa{{ARA\&A}}
\def\apss{{Ap\&SS}}

@@

  We present results from combining a grid-based radiative transfer
  code with a Smoothed Particle Hydrodynamics code to produce a
  flexible system for modelling radiation hydrodynamics. We use a
  benchmark model of a circumstellar disc to determine a robust method
  for constructing a gridded density density distribution from SPH
  particles. The benchmark disc is then used to determine the accuracy
  of the radiative transfer results. We find that the SED and 
  the temperature distribution within the disc are sensitive
  to the representation of the disc inner edge, which depends
  critically on both the grid and SPH resolution. The code is then
  used to model a circumstellar disc around a T-Tauri star. As the
  disc adjusts towards equilibrium vertical motions in the disc are
  induced resulting in scale height enhancements which intercept
  radiation from the central star. Vertical transport of radiation
  enables these perturbations to influence the mid-plane temperature
  of the disc. The vertical motions decay over time and the disc
  ultimately reaches a state of simultaneous hydrostatic and radiative
  equilibrium.

@@

Accurately coupling the radiative transfer component to the SPH
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
circumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

The accuracy of the SPH method has previously been examined in detail
(e.g. \cite{price_2008b} and references therein) so we do not attempt
to benchmark the SPH component of the code here. Likewise the  {\small{TORUS}}
radiative transfer code has also been benchmarked elsewhere
\citep{pinte_2009}. Consequently we can be confident
in the accuracy of the SPH and radiative transfer codes individually so
we focus here on the accuracy of the mechanism by which they are
coupled. Hence the benchmark case presented here is a ``static''
benchmark in which we represent an analytical density distribution
with SPH particles and test whether the radiative transfer code can
calculate accurate results using this SPH density representation.

@@

An important step in setting up grid-based radiative transfer using a
density distribution derived from SPH particles is determining an
effective transformation from the Lagrangian particle description to
the Eulerian grid description.  The representation on grid should
preserve important properties of the SPH model (e.g. total mass) but
also needs to take account of the possibility that a radiative
transfer calculation requires the highest resolution in different
regions to a hydrodynamics calculation (e.g. to resolve opacity
gradients). This section determines a grid construction method which
allows a good degree of control over the number of cells in the grid,
and the location of the highest resolution regions, while maintaining
an accurate representation of the total disc mass.

@@

The density in a given grid cell is calculated using an exponential
kernel smooth. If the sum of the kernel weights is greater than 0.3
the density is normalised by the sum of the kernel weights, to ensure a
smooth density distribution within the interior of the disc. If the
sum of the weights is less than 0.3 then the density is not normalised
by the sum of the weights, in order to avoid numerical effects at the
free surface, as described by \cite{price_2007}. For more details of
the algorithm used to generate the AMR grid from SPH particles see
\cite{rundle_2009}.

@@

For each splitting condition a number of AMR grids were generated
using different values of the mass per cell limit or
$f_{\rm{split}}$. This was repeated for discs represented by $10^5$,
$10^6$ and $10^7$ SPH particles. Figure~\ref{fig:grid_tests} plots the
number of octals generated as a function of mass per cell or
$f_{\rm{split}}$ (solid line), and the percentage error in the disc
mass (dashed line). The percentage error in disc mass is calculated by
comparing the total mass on the AMR grid with the known mass of
$0.011~\rm{M}_{\sun}$ based on the analytical form of the density
distribution, and a positive value indicates that mass on the grid is
too large.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for Cartesian AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:grid_tests}
\end{figure*}

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but it needs to be modified to take into account the increase in 
cell volume as cylindrical polar radius increases, for fixed values of
$dr$ and $dz$.  
%
Consider a Cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in the Cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{az}}}
\end{equation}
if there are $n_{\rm{az}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{az}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in Cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{az}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a Cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

The grid generation tests were run using a cylindrical polar geometry
with the modified mass per cell limit described above and an
unmodified density contrast condition. The results are plotted in
Fig.~\ref{fig:cp_grid_tests}.
\begin{figure*}
  \subfigure[Mass per cell condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set1}}
  \subfigure[Mass per cell condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set1}}
  \subfigure[Mass per cell condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set1}}
  \subfigure[Density contrast condition and $10^5$ particles]{\includegraphics[scale=0.3]{figures/1e5_cyl_set2}}
  \subfigure[Density contrast condition and $10^6$ particles]{\includegraphics[scale=0.3]{figures/1e6_cyl_set2}}
  \subfigure[Density contrast condition and $10^7$ particles]{\includegraphics[scale=0.3]{figures/1e7_cyl_set2}}
  \caption{Number of octals (solid line) and percentage mass error
    (dashed line) for cylindrical polar AMR grids generated using different grid
    splitting methods and different numbers of particles. }
  \label{fig:cp_grid_tests}
\end{figure*}
Using a cylindrical polar geometry results in a significantly reduced
number of octals in the AMR grid but with larger mass errors. Provided
the mass per cell limit is no more than $10^{27}\rm{g}$ then the total
mass is well represented with significantly fewer cells than in a
Cartesian grid. The density contrast condition alone does not
represent the total mass to within 3\% accuracy even with
$f_{\rm{split}}=0.1$. Although this condition successfully adds
enhanced resolution in regions of higher density contrast it needs to
be combined with a mass per cell limit to ensure a robust total mass
representation.

@@

Having investigated the procedure for setting up the radiative transfer
grid we now proceed to studying the accuracy of the results from the
radiative transfer calculation. This was carried out by comparing
temperature distributions and spectral energy distributions (SEDs)
from the benchmark disc. Temperature distribution errors provide a
clear picture of where errors in the radiative transfer calculation
are located whereas SEDs provide a direct link to an observable
quantity and are a stringent test of the accuracy of the radiative
transfer calculation.

The disc was initially modelled using $10^5$ SPH particles with an AMR
grid constructed from a mass per cell limit of $5\times10^{26}$g.  The
mass per cell limit was chosen in preference to the density contrast
limit to give the highest resolution near the central region of the
disc, which we expect to be most important for determining the
SED. SEDs were calculated using the {\small{TORUS}} code for viewing
angles of 12.5 and 77.5 degrees (where an inclination angle of zero
degrees indicates that the disc is viewed face-on) and the results are
plotted in Fig.~\ref{fig:seds} (dashed line). Also plotted in
Fig.~\ref{fig:seds} are the benchmark results of \cite{pascucci_2004}
(solid line).
\begin{figure*}
  \includegraphics[scale=0.3]{figures/sed_013}
  \includegraphics[scale=0.3]{figures/sed_077}
  \caption{SEDs from the benchmark disc with $10^5 $ particles. SEDs
    are shown for inclination angles of 12.5 degrees (left) and 77.5
    degrees (right). The benchmark SED is plotted as a solid line and
    the {\small{TORUS}} SED is plotted as a dashed line.}
  \label{fig:seds}
\end{figure*}
Both SEDs show significant departures from the benchmark results;
there is excess emission between 10-100 $\mu\rm{m}$ and at a viewing
angle of 77.5 degrees there is also a significant reduction in flux
below 1~$\mu\rm{m}$. The density distribution in the central region of
this disc is plotted in Fig.~\ref{fig:basic_dens} which shows that the
central 1AU gap is not represented. There are two effects in
operation; firstly the grid resolution in this region is larger than
1AU and secondly the smoothing lengths of the particles are larger
than 1AU. In order to correctly represent the central gap there needs
to be higher grid resolution around the central source and more SPH
particles are required so that the smoothing lengths are smaller.
\begin{figure}
\includegraphics[scale=0.2]{figures/basic_dens}
\caption{Density in central region of disc with $10^5$ particles. Axis
units are $10^{13}$cm.}
\label{fig:basic_dens}
\end{figure}

@@

The disc is then evolved using the combined SPH and radiative transfer
code. In this configuration the internal energies of the SPH particles
are set using the temperatures calculated by the radiative transfer
code. The SPH part of the code does not change the particle
temperatures and they retain the temperature assigned by the radiative
transfer code until the next radiation time step. The temperatures
calculated by the radiative transfer code are equilibrium
temperatures, as it is assumed that the time scale for radiative
equilibrium is short compared to the dynamical time scale of the disc
(i.e. the ``prompt escape'' regime described by \cite{nayakshin_2009}).
A radiative transfer calculation is performed after every four
hydrodynamics time steps (equivalent to every 2.4 years). The frequency
with which the radiative transfer is calculated is determined by the
sound crossing time in the inner regions of the disc as this is the
time scale over which the scale height in this region will vary and
effect the transfer of radiation into the outer parts of the disc.

A cylindrical polar grid, with adaptive $r$ and $z$ cells and 16
azimuthal cells, was used for the radiative transfer step. The AMR grid
was constructed using a maximum mass per cell of
$3.0\times10^{25}~\rm{g}$ and a density splitting condition of
$f_{\rm{split}}=0.1$. The density in a cylindrical region around the
source was reduced to a very low value to ensure that the disc did not
completely obscure the source. A real circumstellar disc will be
truncated by the stellar magnetosphere so there is a physical
justification for enforcing a low density region around the star. The
radius of this region is $R_{\rm{gap}}=1.0~\rm{AU}$ which corresponds
to the accretion radius of the SPH sink particle. Two extra levels of
grid refinement were added in a box around the central source to
ensure that the grid had enough resolution to allow the gap around the
source to be resolved. The first box of extra refinement has size
$10^{14}~\rm{cm}$ with cells no larger than $10^{13}~\rm{cm}$ and the
second box has size $4\times10^{13}~\rm{cm}$ with cells no larger than
$4\times10^{12}~\rm{cm}$.

@@

Plots of the disc internal energy are shown in Fig.~\ref{fig:log_u} at
six different times during the simulation. Figure~\ref{fig:frame1}
shows the initial state before the radiative transfer is switched on
when there is a vertically isothermal equation of state. The second
frame (Fig.~\ref{fig:frame2} is shortly after the radiative transfer
has been switched on. The inner disc has collapsed and there is
heating seen in parts of the disc exposed to radiation from the
central star. In Fig.~\ref{fig:frame3} a scale height enhancement is
present in the disc at a radius of approximately
$3\times10^{15}~\rm{cm}$ which shadows the region immediately behind
causing it to cool. Figure~\ref{fig:frame4} shows several such scale
height enhancements and in each case there is cooling behind the
enhancement as the scale height fluctuations are able to affect the
mid-plane temperature due to vertical transport of radiation. In
Fig.~\ref{fig:frame5} the inner region of the disc has settled and the
scale height enhancements are seen only in the outer disc and by the
last frame (Fig.~\ref{fig:frame6}) the disc has reached a steady state
in hydrostatic and radiative equilibrium.
\begin{figure*}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6675}\label{fig:frame1}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6691}\label{fig:frame2}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6763}\label{fig:frame3}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC7363}\label{fig:frame4}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC9363}\label{fig:frame5}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_run2_DSC1024}\label{fig:frame6}}
  \caption{Particle plots showing internal energy in cylindrical
    polar co-ordinates. Axis units are cm.}
  \label{fig:log_u}
\end{figure*}

Transient oscillations, as the disc adjusts towards equilibrium, are
observed whether radiative transfer is included or not. They are
vertical motions which decay more rapidly at smaller radii and hence
give the appearance of propagating outwards, even though the motion is
vertical. Other disc models using radiation hydrodynamics have found
oscillatory solutions with inwards propagating waves
(e.g. \cite{min_2009}) which are fundamentally different to the
transient effects seen in this case. However it should be noted that
\cite{min_2009} only found an oscillatory solution in their most
optically thick disc, which is significantly more massive than our
model disc, and may represent a different regime to that considered
here.

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SEDs. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a manageable
level with a realistic number of SPH particles in the disc.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. As the disc settles towards an equilibrium state,
vertical motions are seen which give the appearance of an outward
propagating scale height fluctuation. This is a vertical motion which
decays more rapidly at smaller radii and is seen whether radiative
transfer is included or not. These are not the same as the
instabilities seen in other radiation hydrodynamics disc models
\citep{min_2009}. One consequence of these fluctuations is that they
can intercept radiation from the central star which can then be
vertically transported into the mid-plane of the disc hence even
thought the fluctuations are of a small amplitude they exert an
influence on the mid-plane temperature.
